# Multiple-linear regression.

**Created by Robin Cunningham, UNC Chapel Hill**

**Modified by Jan Hannig, UNC Chapel Hill**

***HOMEWORK 8 - Predicting Baby-weight***
<br><br>
A Multi-linear regression analysis. **Please compose all answers in this R-Markdown document**.
<br><br>
1. A file containing data on 1236 live births can be found at
'https://drive.google.com/open?id=0B2lwGKhIFjYYbDY3eWVubEZzX28'. We will use this dataset to construct a multi-linear model for predicting birthweight from other variables.
<br><br>
a. Write and execute code to read the csv file 'babies.csv', assign it to the object 'babies' and summarize the variables.
```{r, eval=TRUE}
babies = read.csv("babies.csv")
summary(babies)

```
<br>
b. As you can see, there are 8 variables and a fair number of missing data points. Remove all cases with missing data and assign the resulting data frame to 'bbycomp'.
```{r, eval=TRUE}
bbycomp = na.omit(babies)
summary(bbycomp)
```
<br>
c. A **Dataset Codebook** is a guide to what each of the variables represents. Note that for the purpose of this study, we will consider each variable to be numerical. Complete the comment box below to create a codebook for these data. Include units if you can figure them out.
```
Variables:
i. Case - Which mother-baby pairing is being described
ii. bwt - Weight of mbaby in ounces
iii. gestation - Number of days mother was pregnant
iv. parity - Has mother been pregnant before
v. age - Age of Mother at babies birth
vi. height - Height of baby in cm
vii. weight - Weight of mother
viii. smoke - indicator of whether or not the mother smoked

```
<br>
d.Do some exploratory analysis by looking at histograms of the 7 variables and plots of bwt versus each of the six explanatory variables. In the comment box below, make a note of any concerns. ***It will save some typing to assign the right variables to `Y, X1, ..., X6`, so I did that for you.***
```{r, eval =FALSE}

#Assign Short variable names
Y<- bbycomp$bwt
X1 <- bbycomp$gestation
X2 <- bbycomp$parity
X3 <- bbycomp$age
X4 <- bbycomp$height
X5 <- bbycomp$weight
X6 <- bbycomp$smoke

#Histograms
hist(Y, xlab = "BWT")
hist(X1, xlab = "Gestation")
hist(X2, xlab = "Parity")
hist(X3, xlab = "Age")
hist(X4, xlab = "Height")
hist(X5, xlab = "Weight")
hist(X6, xlab = "Smoke")



#Plots of Y versus X_i

plot(Y, X1, xlab = "BWT", ylab = "Gestation")
plot(Y, X2, xlab = "BWT", ylab = "Parity")
plot(Y, X3, xlab = "BWT", ylab = "Age")
plot(Y, X4, xlab = "BWT", ylab = "Height")
plot(Y, X5, xlab = "BWT", ylab = "Weight")
plot(Y, X6, xlab = "BWT", ylab = "Smoke")

```
```
Comments:
The birth weight appears to be an approximately normal distrobution
The relationship between the variables is hard to approximate. 


```
<br>
e. Run the full model using all of the other variables (besides Case) to explain Birthweight (bwt). Store the model as `full.lm` and create a summary of the model.
```{r, eval=TRUE}
full.lm = lm(bwt~ gestation + parity + age + height + weight + smoke, data = bbycomp)
summary(full.lm)
anova(full.lm)
```
<br>
f. Use the summary to conduct an ANOVA test to see if at least one of the coefficients is significantly different from zero. State the results in the comment box below.
```
All of the values have F-statistics and P values that indicate rejecting the null hypothesis except for age which is not significantly different than 0. 
```
g. Now perform backward elimination in the following manner: First, eliminate the predictor whose removal causes the greatest improvement in adjusted R-squared. Continue in this manner until removing any remaining predictors causes Adjusted R-squared to fall. <br>
Begin by finding the 5-predictor model that increases adjusted R-squared by the most. Include the model and summary in the codebox below.
```{r, eval=TRUE}
mod5 = lm(bwt~ gestation + parity  + height + weight + smoke, data = bbycomp)
summary(mod5)

```
<br>
h. Should we stick with the 6-predictor model or continue? Explain.
```
We should use the 5 predictor model. The ajdusted R squared is higher for the 5 variable model than it is for the 6. All the variables in the new mod5 are significant to the model as well. 
```
<br>
i. Now find the best 4-variable model using the same criterion and include it in the code box below. Include a summary of the model.
```{r, eval=TRUE}
mod4 = lm(bwt~ gestation + parity  + height  + smoke, data = bbycomp)
summary(mod4)
```
<br>
j. According to the Adj. R-squared criterion, should we stick with 5-predictors or continue? Explain.
```
We should stay with the 5 variable model because the adjusted r squared is higher. 
```
***Note: even though our criterion says to stick with 5 predictors, I would seriously consider dropping X5 anyway, because the p-value is very close to 0.05 and we have lots of predictors. (Think about why having lots of predictors matters for this!) Also, the value of Adjusted R-squared is only reduced slightly and a parsimonious model is easier to understand and more robust for predictions***
k. Using the best 5-predictor model that you found, find a 95% confidence interval for the average birthweight among all babies for which (gestation, parity, age, height, weight, smoke) = (290, 1, 22, 60, 110, 0). (One line of code will do it.)
```{r, eval= TRUE}
newbaby = data.frame(gestation = 290, parity = 1, age = 22, height = 60, weight = 110, smoke = 0)
predict(mod5, newbaby, interval = "confidence")
```
<br>
l. Using the best 5-predictor model that you found, find a 95% confidence interval for the birthweight of the next baby for which (gestation, parity, age, height, weight, smoke) = (290, 1, 22, 60, 110, 0). (Again, don't make it hard ... one line.)
```{r, eval= TRUE}
predict(mod5, newbaby, interval = "prediction")
```
<br>
m. In plain English, interpret the coefficients in the least squares model for `height` and `smoke`.
```
If a mother is a smoker, the baby's predicted weight will decrease 8.4 ounces. For every centimeter taller the baby is, it is predicted to be 1.15 ounces heavier.
```
<br><br>
2. Run diagnostics on the final 5-predictor model you selected. Include appropriate residual plots and your comments on the quality and usefulness of the fit. (Make your own codeboxes and comment boxes.)
```{r}
plot(mod5)
```

```
The model appears to fit the data fairly well. Most points lie nicely on the QQ plot line. There appears to be a little non-linear reaction at the tails of the plot. 
```

<br><br>
3. In the plots you created before doing any regressions, there were apparent outliers with regard to both X1 and X4. Without doing the work, say what steps you would take to evaluate whether we should consider removing these outliers. (Your own comment box.)
```
To determine the impact of the outliers, I would create a subset of the data without those points and refit the model. A significant jump in r^2 indicates that the outliers were impactful.
```

<br><br>
4. Using `Forward Addition`, choose a "best" multilinear model for this data set. Begin by choosing the single predictor that gives the highest value of Adjusted R-squared and continue by adding variables until Adjusted R-squared has been maximized. ***Your answer should consist of a set of nested models with increasing Adjusted R-squared***
```{r}
mod1 = lm(bwt~gestation, data = bbycomp)
mod2 = lm(bwt~gestation+smoke, data = bbycomp)
mod3 = lm(bwt~gestation+smoke+height, data = bbycomp)
mod4 = lm(bwt~gestation+smoke+height+parity, data = bbycomp)
summary(mod1)
summary(mod2)
summary(mod3)
summary(mod4)
summary(mod5)
```

<br><br>
5. For the sequence of nested models above, conduct an ANOVA test comparing each model to the previous, reduced model to see if the new coefficient is statistically different from zero compared to the reduced model. Show the code for each test and state the results.
```{r}
anova(mod1, mod2)
anova(mod2, mod3)
anova(mod3, mod4)
anova(mod4, mod5)
```

```
The tests are all significant at the .05 level. They are progressively less significant as more variables are added.
```

<br><br>
6. Would considering some interactions make sense? Try to add some interactions to the best models and see what happens.
```{r}
mymod = lm(bwt~gestation + (height*smoke), data = bbycomp)
summary(mymod)
```

```
I created an interaction between smoke and height and it cuased smoke and the joint term to be insignificant. Height and gestation alone are still significant but the r squared value is lower than mod5 seen previously indicating that interaction will not help the model. 
```
