# Matrices in R
**Created by John Palowitch, UNC Chapel Hill**

**Edited by Jan Hannig, UNC Chapel Hill**


### Conceptual Questions 
Please read the document ***Matrices in R.docx*** before completing this assignment. This assignment has two goals:

-Learning new commands in R from a written source.

-Using R for Matrix Calculations
<br><br>
1. Suppose:
\[
\mathbf{A} = \begin{bmatrix}2&3&4&5\\-1&0&3&2\end{bmatrix},\;\;\;\;\mathbf{B} = \begin{bmatrix}4&-1&10&0\\-2&-2&5&6\end{bmatrix}
\]
\[
\mathbf{a} = \begin{bmatrix}6&2&-11&1\end{bmatrix},\;\;\;\;\mathbf{b} = \begin{bmatrix}4\\2\\3\\0\end{bmatrix},\;\;\;\;\mathbf{c} = \begin{bmatrix}-3\\-10\end{bmatrix}
\]
    a. Write down the dimensions of the following (you can use a comment box that you create). If the operation is not possible, write "N.P."
        i. $\mathbf{A}^T\mathbf{B}$   4X4 (first one done for you!)
        ii. $\mathbf{B}\mathbf{A}^T$  4X4
```{r}
A = rbind(c(2:5), c(-1, 0, 3, 2))
B = rbind(c(4, -1, 10, 0), c(-2, -2, 5, 6))
a = cbind(6,2,-11, 1)
b = rbind(4, 2, 3, 0)
c = rbind(-3, -10)
dim(t(A)%*%B)

```
        iii. $\mathbf{A}\mathbf{B}$  N.P.
```{r}
#dim(A%*%B)
```
        iv. $\mathbf{A}^T\mathbf{A}$ 4X4
```{r}
dim(t(A)%*%A)
```
        v. $\mathbf{b}^T\mathbf{A}$ N.P.
```{r}
#dim(t(b)%*%A)
```
        vi. $\mathbf{a}\mathbf{A}^T$ 1X2
```{r}
dim(a%*%t(A))
```
        vii. $\mathbf{B}\mathbf{b}$ 2X1
```{r}
dim(B%*%b)
```
        viii. $\mathbf{B}\mathbf{c}^T$ N.P.
```{r}
#dim(B%*%t(c))
```
        ix. $\mathbf{a}\mathbf{c}$ N.P.
```{r}
#dim(a%*%c)
```
        x. $\mathbf{b}\mathbf{a}$ 4X4
```{r}
dim(b%*%a)
```
    b. Again, write down the dimensions of the following (on this sheet, next to the expression). If the operation is not possible, write "N.P."
        i. $(\mathbf{A}^T\mathbf{B})^{-1}\mathbf{a}$ N.P.
```{r}
step1 = (t(A)%*%B)
det(step1)
```
        ii. $\mathbf{c}(\mathbf{B}\mathbf{A}^T)^{-1}$ N.P.
```{r}
step1 = solve(B%*%t(A))
#dim(c%*%step1)
```
        iii. $(\mathbf{B}^T\mathbf{B})^{-1}\mathbf{B}^T\mathbf{c}$ N.P.
```{r}
#step1 = solve(t(B)%*%B)
#dim(step1%*%t(B)%*%c)
```        
        iv. $\mathbf{b}^T\mathbf{A}^T\mathbf{c}$ 1X1
```{r}
dim(t(b)%*%t(A)%*%c)
```
        v. $(\mathbf{B}\mathbf{B}^T)^{-1}\mathbf{B}\mathbf{b}$ 2X1
```{r}
step1 = solve(B%*%t(B))
dim(step1%*%B%*%b)
```
      
        
### Applied Questions 
2. In a code block below, create and print the following matrices and vectors:
\[
\mathbf{A} = \begin{bmatrix}-6&0&1\\1&10&7.5\end{bmatrix},\;\;\;\;\mathbf{B} = \begin{bmatrix}-15&1&1\\5.25&1&2.5\end{bmatrix}
\]
\[
\mathbf{a} = \begin{bmatrix}6&2&1\end{bmatrix},\;\;\;\;\mathbf{b} = \begin{bmatrix}2\\-7\end{bmatrix}
\]

```{r}
A = rbind(c(-6,0,1),c(1,10,7.5))
B = rbind(c(-15,1,1), c(5.25,1,2.5))
a = cbind(6,2,1)
b = rbind(2,-7)
A
B
a
b

```

3. Compute (with **R**, in a new code block below) and print the following products:
    a. $\mathbf{A}^T\mathbf{B}$
```{r}
t(A)%*%B
```
    b. $\mathbf{B}\mathbf{a}^T$
```{r}
B%*%t(a)
```
    c. $\mathbf{b}^T\mathbf{A}$
```{r}
t(b)%*%A
```
    d. $\mathbf{A}^T\mathbf{b}$
```{r}
t(A)%*%b
```
    e. $\mathbf{a}^T\mathbf{a}$
```{r}
t(a)%*%a
```
    f. $\mathbf{a}\mathbf{a}^T$
```{r}
a%*%t(a)
```
    g. $(\mathbf{A}^T\mathbf{A})^{-1}$
```{r}
#solve(t(A)%*%A)
```

The determinant of this matrix is 0, therefore it is not invertable.

    h. $(\mathbf{A}^T\mathbf{A})^{-1}\mathbf{A}^T\mathbf{b}$
```{r}
#solve(t(A)%*%A)%*%A%*%b
```

The determinant of this matrix is 0, therefore it is not invertable.

    i. $\mathbf{b}^T\mathbf{A}\mathbf{a}^T$
```{r}
t(b)%*%A%*%t(a)
```
    j. $\mathbf{a}\mathbf{B}\mathbf{A}^T\mathbf{b}$
```{r}
#a%*%B%*%t(A)%*%b
```

a is a 1X3 and B is a 1X2, the dimensions do not line up for matrix multiplication.


